---
title: "Lab 2: Random variables and probability distributions"
output: html_notebook
---

https://campus.datacamp.com/courses/computational-finance-and-financial-econometrics-with-r/lab-2-random-variables-and-probability-distributions?ex=1

##Compute probabilities

In this lab you will become more familiar with random variables and probability distributions.

Suppose $X$ is a normally distributed random variable with a mean $0.05$ and a variance $(0.10)^2$. Start by computing some probabilities.

Probabilities for normally distributed random variables can be computed with the function <code>pnorm()</code>. For instance, if $Z$ is a standard normal random variable, $P_r(Z\leqslant0.5)$ can be computed with <code>pnorm(0.5)</code>. You can supply the mean and the standard deviation of the normal distribution via the arguments <code>mean</code> and <code>sd</code>, respectively.

```{r}
#X ~ N(0.05, (0.10)^2)
mu_x <-0.05
sigma_x <- 0.10

# Pr(X > 0.10)
# 1 - Pr(X < 0.10)
1 - pnorm(0.1, mean=mu_x, sd=sigma_x)

# Pr(X < -0.10)
pnorm(-0.1, mean=mu_x, sd=sigma_x)

# Pr(-0.05 < X < 0.15)
# Pr(-0.05 < X < 0.15) = Pr(X < 0.15) - Pr(X < -0.05)
pnorm(0.15, mean=mu_x, sd=sigma_x) - pnorm(-0.05, mean=mu_x, sd=sigma_x)
```

##Compute quantiles

Continue with computing quantiles of the normally distributed random variable $X$ with a mean $0.05$ and a variance $(0.10)^2%.

Quantiles of normally distributed random variables can be computed with the function <code>qnorm()</code>. For instance, if $Z$ is a standard normal random variable, the 10% quantile can be computed with <code>qnorm(0.1)</code>. You can supply the mean and the standard deviation of the normal distribution via the arguments <code>mean</code> and <code>sd</code>, respectively.

```{r}
# 1%, 5%, 95% and 99% quantile
qnorm(c(0.01, 0.05, 0.95, 0.99), mean=mu_x, sd=sigma_x)
```

##Compute densities

Now consider two stocks: Microsoft (ticker: MSFT) and Starbucks (ticker: SBUX). Let $X$ denote the monthly return on the Microsoft stock and let $Y$ denote the monthly return on the Starbucks stock. Assume that $X$ is normally distributed with a mean $0.05$ and a variance $(0.10)^2$, and that $Y$ is normally distributed with a mean $0.025$ and a variance $(0.05)^2$.

Values of a normal density can be computed with function <code>dnorm()</code>. For instance, the value of the standard normal density at 0 can be computed with <code>dnorm(0)</code>. You can supply the mean and the standard deviation of the normal distribution via the arguments <code>mean</code> and <code>sd</code>, respectively.

```{r}
# Normally distributed monthly returns
x_vals <- seq(-0.25, 0.35, length.out = 100)
MSFT <- dnorm(x_vals, mean=0.05, sd=0.10)
SBUX <- dnorm(x_vals, mean=0.025, sd=0.05)
```

##Plot normal curve

```{r}
plot(x_vals, MSFT)
plot(x_vals, MSFT, col = 'blue', type = 'l', ylab = 'Normal curves', ylim = c(0,8))
```

##Add second normal curve

```{r}
# Normal curve for MSFT
plot(x_vals, MSFT, type = "l", col = "blue", ylab = "Normal curves", ylim = c(0, 8))

# Add a normal curve for SBUX
lines(x_vals, SBUX, col = "red")

# Add a plot legend
legend("topleft", legend = c("Microsoft", "Starbucks"), 
       col = c("blue", "red"), lty = 1)
```

##Determine the value-at-risk of simple monthly returns

Consider again the Microsoft stock. Assume that the simple monthly return $R$ is normally distributed with a mean $0.04$ and a variance $(0.09)^2$. The initial wealth to be invested over the month is $100,000.

Determine the 1% and the 5% value-at-risk (VaR) over the month on the investment. That is, determine the loss in investment value that may occur over the next month with a 1% probability and with a 5% probability.

```{r}
# R ~ N(0.04, (0.09)^2) 
mu_R <- 0.04
sigma_R <- 0.09

# Initial wealth W0 equals $100,000
W0 <- 100000

# The 1% value-at-risk
W0 * qnorm(0.01, mean=mu_R, sd=sigma_R)

# The 5% value-at-risk
W0 * qnorm(0.05, mean=mu_R, sd=sigma_R)
```

##Determine the value-at-risk of continuously compounded monthly returns

Instead of the simple monthly return, now look at the continuously compounded monthly return rr of the Microsoft stock. Assume that rr is normally distributed with a mean $0.04$ and a variance $(0.09)^2$. The initial wealth to be invested over the month is $100,000.

Determine the 1% and the 5% value-at-risk (VaR) over the month on the investment. That is, determine the loss in investment value that may occur over the next month with a 1% probability and with a 5% probability.

Use the fact that the continuously compounded return quantile can be transformed to a simple return quantile with the transformation $R=e^r-1$. The exponential $e^r$ can easily be computed with <code>exp(r)</code>.

```{r}
# r ~ N(0.04, (0.09)^2) 
mu_r <- 0.04
sigma_r <- 0.09

# Initial wealth W0 equals $100,000
W0 <- 100000

# The 1% value-at-risk
W0 * (exp(qnorm(0.01, mean = mu_r, sd = sigma_r)) - 1)

# The 5% value-at-risk
W0 * (exp(qnorm(0.05, mean = mu_r, sd = sigma_r)) - 1)
```

##Compute simple monthly returns

For the remaining exercises, consider a one-month investment in two Northwest stocks: Amazon and Costco.

Suppose you buy Amazon and Costco at the end of September at $P_{A,t-1}=38.23$, $P_{C,t-1}=41.11$ and then sell at the end of October for $P_{A,t}=41.29$, $P_{C,t}=41.74$. Compute the simple monthly returns for the two stocks.

Note that the above prices were the actual closing prices for 2004 taken from Yahoo!.

```{r}
# Vectors of prices
PA <- c(41.29, 38.23)
PC <- c(41.74, 41.11)

# Simple monthly returns
RA <- (41.29- 38.23)/38.23
RC <- (41.74- 41.11)/41.11
```
